---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
<!-- Variables marked with double curly braces will be transformed beforehand: -->
<!-- `packagename` is extracted from the DESCRIPTION file -->
<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# _simulist_: Simulate line list data <img src="man/figures/logo.svg" align="right" width="120" />

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/license/mit/)
[![R-CMD-check](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/{{ gh_repo }}/branch/main/graph/badge.svg)](https://app.codecov.io/gh/{{ gh_repo }}?branch=main)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.10471459.svg)](https://doi.org/10.5281/zenodo.10471459)
<!-- badges: end -->

`{simulist}` is an R package to simulate individual-level infectious disease outbreak data, including line lists and contact tracing data. It can often be useful to have synthetic datasets like these available when demonstrating outbreak analytics techniques or testing new analysis methods.

`{simulist}` is developed at the [Centre for the Mathematical Modelling of Infectious Diseases](https://www.lshtm.ac.uk/research/centres/centre-mathematical-modelling-infectious-diseases) at the [London School of Hygiene and Tropical Medicine](https://www.lshtm.ac.uk/) as part of [Epiverse-TRACE](https://data.org/initiatives/epiverse/).

## Installation

You can install the development version of `{simulist}` from
[GitHub](https://github.com/) with:

``` r
# check whether {pak} is installed
if(!require("pak")) install.packages("pak")
pak::pak("epiverse-trace/simulist")
```

## Quick start

```{r load-simulist}
library(simulist)
library(epiparameter)
```

The line list simulation requires that we define a contact interval, onset-to-hospitalisation delay, and onset-to-death delay. We can load these from the library of epidemiological parameters in the `{epiparameter}` R package if available, or if these are not in the database yet (such as the contact interval for COVID-19) we can define them ourselves.

```{r create-epidists}
# create COVID-19 contact interval
contact_interval <- epiparameter::epidist(
  disease = "COVID-19",
  epi_dist = "contact interval",
  prob_distribution = "gamma",
  prob_distribution_params = c(shape = 1, scale = 1)
)

# get onset to hospital admission from {epiparameter} database
onset_to_hosp <- epiparameter::epidist_db(
  disease = "COVID-19",
  epi_dist = "onset to hospitalisation",
  single_epidist = TRUE
)

# get onset to death from {epiparameter} database
onset_to_death <- epiparameter::epidist_db(
  disease = "COVID-19",
  epi_dist = "onset to death",
  single_epidist = TRUE
)
```

To simulate a line list for COVID-19 with an assumed average number of contacts of 2 and a probability of infection per contact of 0.5, we use the `sim_linelist()` function. The mean number of contacts and probability of infection determine the outbreak reproduction number, if the resulting reproduction number is around one it means we will likely get a reasonably sized outbreak (10 - 1000 cases, varying due to the stochastic simulation). _Take care when setting the mean number of contacts and the probability of infection, as this can lead to the outbreak becoming extremely large_.

```{r sim-linelist}
set.seed(1)
linelist <- sim_linelist(
  mean_contacts = 2,
  contact_interval = contact_interval,
  prob_infect = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death
)
head(linelist)
```

In this example, the line list is simulated using the default values (see `?sim_linelist`). The default hospitalisation risk is assumed to be 0.2 (i.e. there is a 20% probability an infected individual becomes hospitalised) and the start date of the outbreak is 1st January 2023. To modify either of these, we can specify them in the function.

```{r sim-linelist-diff-args}
linelist <- sim_linelist(
  mean_contacts = 2,
  contact_interval = contact_interval,
  prob_infect = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death,
  hosp_risk = 0.01,
  outbreak_start_date = as.Date("2019-12-01")
)
head(linelist)
```

To simulate a table of contacts of cases (i.e. to reflect a contact tracing dataset) we can use the same parameters defined for the example above.

```{r, sim-contacts}
contacts <- sim_contacts(
  mean_contacts = 2, 
  contact_interval = contact_interval, 
  prob_infect = 0.5
)
head(contacts)
```

If both the line list and contacts table are required, they can be jointly simulated using the `sim_outbreak()` function. This uses the same inputs as `sim_linelist()` and `sim_contacts()` to produce a line list and contacts table of the same outbreak (the arguments also have the same default settings as the other functions).

```{r, sim-outbreak}
outbreak <- sim_outbreak(
  mean_contacts = 2,
  contact_interval = contact_interval,
  prob_infect = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death
)
head(outbreak$linelist)
head(outbreak$contacts)
```

## Help 

To report a bug please open an [issue](https://github.com/epiverse-trace/simulist/issues/new/choose).

## Contribute 

Contributions to `{simulist}` are welcomed. Please follow the [package contributing guide](https://github.com/epiverse-trace/.github/blob/main/CONTRIBUTING.md).

## Code of Conduct

Please note that the `{simulist}` project is released with a 
[Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

## Citing this package

```{r message=FALSE, warning=FALSE}
citation("simulist")
```
