---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
<!-- Variables marked with double curly braces will be transformed beforehand: -->
<!-- `packagename` is extracted from the DESCRIPTION file -->
<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# _simulist_: Simulate line list data <img src="man/figures/logo.svg" align="right" width="120" />

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/license/mit/)
[![R-CMD-check](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/{{ gh_repo }}/branch/main/graph/badge.svg)](https://app.codecov.io/gh/{{ gh_repo }}?branch=main)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`{simulist}` is an R package to simulate infectious disease outbreak data, including line lists and contacts.

`{simulist}` is developed at the [Centre for the Mathematical Modelling of Infectious Diseases](https://www.lshtm.ac.uk/research/centres/centre-mathematical-modelling-infectious-diseases) at the [London School of Hygiene and Tropical Medicine](https://www.lshtm.ac.uk/) as part of [Epiverse-TRACE](https://data.org/initiatives/epiverse/).

## Installation

You can install the development version of `{simulist}` from
[GitHub](https://github.com/) with:

``` r
# check whether {pak} is installed
if(!require("pak")) install.packages("pak")
pak::pak("epiverse-trace/simulist")
```

## Quick start

```{r load-simulist}
library(simulist)
library(epiparameter)
```

The line list simulation requires a serial interval, onset-to-hospitalisation delay, and onset-to-death delay. We can load these load these from the library of epidemiological parameters in the `{epiparameter}` R package, or if these are not available, such as the serial interval for COVID-19 we can create them ourselves.

```{r create-epidists}
# create COVID-19 serial interval
serial_interval <- epiparameter::epidist(
  disease = "COVID-19",
  epi_dist = "serial interval",
  prob_distribution = "gamma",
  prob_distribution_params = c(shape = 1, scale = 1)
)

# get onset to hospital admission from {epiparameter} database
onset_to_hosp <- epiparameter::epidist_db(
  disease = "COVID-19",
  epi_dist = "onset to hospitalisation",
  single_epidist = TRUE
)

# get onset to death from {epiparameter} database
onset_to_death <- epiparameter::epidist_db(
  disease = "COVID-19",
  epi_dist = "onset to death",
  single_epidist = TRUE
)
```

To simulate a line list for COVID-19 with an assumed reproduction number (`R`) of 1.1 we use the `sim_linelist()` function.

```{r sim-linelist}
linelist <- sim_linelist(
  R = 1.1,
  serial_interval = serial_interval,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death
)
head(linelist)
```

In this example, the line list is simulated using the default values (see `?sim_linelist`). The default hospitalisation rate is 0.2 (or 20% of individual infected become hospitalised) and the start date of the outbreak is 1st January 2023. To modify either of these to make them more realistic we can specify them.

```{r sim-linelist-diff-args}
linelist <- sim_linelist(
  R = 1.1,
  serial_interval = serial_interval,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death,
  hosp_rate = 0.01,
  outbreak_start_date = as.Date("2019-12-01")
)
head(linelist)
```

To simulate a table of contacts we can reuse the serial interval from the example above, and we additionally need a contact distribution. This distribution represents the variability in number of contacts that each person in the population has.

```{r, sim-contacts}
contact_distribution <- epiparameter::epidist(
  disease = "COVID-19",
  epi_dist = "contact_distribution",
  prob_distribution = "pois",
  prob_distribution_params = c(l = 5)
)

contacts <- sim_contacts(
  R = 1.1,
  serial_interval = serial_interval,
  contact_distribution = contact_distribution
)
head(contacts)
```

If both the line list and contacts table are required, they can be jointly simulated using the `sim_outbreak()` function. This uses the same inputs as `sim_linelist()` and `sim_contacts()` to produce a line list and contacts table of the same outbreak (the arguments also have the same default settings as the other functions).

```{r, sim-outbreak}
outbreak <- sim_outbreak(
  R = 1.1, 
  serial_interval = serial_interval, 
  onset_to_hosp = onset_to_hosp, 
  onset_to_death = onset_to_death, 
  contact_distribution = contact_distribution
)
head(outbreak$linelist)
head(outbreak$contacts)
```

## Help 

To report a bug please open an [issue](https://github.com/epiverse-trace/simulist/issues/new/choose).

## Contribute 

Contributions to `{simulist}` are welcomed. Please follow the [package contributing guide](https://github.com/epiverse-trace/.github/blob/main/CONTRIBUTING.md).

## Code of Conduct

Please note that the `{simulist}` project is released with a 
[Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

## Citing this package

```{r message=FALSE, warning=FALSE}
citation("simulist")
```
