---
title: "Getting Started with {simulist}"
output: 
  bookdown::html_vignette2:
    code_folding: show
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Getting Started with {simulist}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is an introductory vignette to the {simulist} R package. {simulist} simulates two types of common epidemiological data collected during infectious disease outbreaks: 1) a line list, which provides individual-level descriptions of cases in an outbreak; 2) a contact dataset, which provides details of which others individuals were in contact with each of the cases.

The main function in the {simulist} package is `sim_linelist()`. This functions takes in arguments that control the dynamics of the outbreak, such as the contact interval (i.e. time delay between contacts), and outputs a line list table (`<data.frame>`) with case information for each infected individual.

For this introduction we will simulate a line list for the early stages of a COVID-19 (SARS-CoV-2) outbreak. This will require two R packages: {simulist}, to produce the line list, and {epiparameter} to provide epidemiological parameters, such as onset-to-death delays.

```{r setup}
library(simulist)
library(epiparameter)
```

First we load in some data that is required for the line list simulation. Data on epidemiological parameters and distributions are read from the {epiparameter} R package.

```{r read-epidist}
# create contact interval (not available from {epiparameter} database)
contact_interval <- epidist(
  disease = "COVID-19",
  epi_dist = "contact interval",
  prob_distribution = "gamma",
  prob_distribution_params = c(shape = 1, scale = 1)
)

# get onset to hospital admission from {epiparameter} database
onset_to_hosp <- epidist_db(
  disease = "COVID-19",
  epi_dist = "onset to hospitalisation",
  single_epidist = TRUE
)

# get onset to death from {epiparameter} database
onset_to_death <- epidist_db(
  disease = "COVID-19",
  epi_dist = "onset to death",
  single_epidist = TRUE
)
```

The seed is set to ensure the output of the vignette is consistent. When using {simulist}, setting the seed is not required unless you need to simulate the same line list multiple times.

```{r, set-seed}
set.seed(123)
```

The mean (average) number of contacts (`mean_contacts`) is the first argument in `sim_linelist()` and with the contact interval and probability of infection per contact (`prob_infect`) will control the growth rate of the simulated epidemic. Here we set the mean number of contacts as 2 and the probability of infection as 0.5 (on average half of contacts become infected). The minimum requirements to simulate a line list are the mean number of contacts (`mean_contacts`), the serial interval, probability of infection, onset-to-hospitalisation delay and onset-to-death delay.

```{r, sim-linelist}
linelist <- sim_linelist(
  mean_contacts = 2,
  contact_interval = contact_interval,
  prob_infect = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death
)
head(linelist)
```

## Case type 

During an infectious disease outbreak it may not be possible to confirm every infection as a case. A confirmed case is typically defined via a diagnostic test. There are several reasons why a case may not be confirmed, including limited testing capacity and mild or non-specific early symptoms, especially in fast growing epidemics. We therefore include two other categories for cases: probable and suspected. For example, probable cases may those that show clinical evidence for the disease but have not, or cannot, be confirmed by a diagnostic test. Suspected cases are those that are possibly infected but do not show clear clinical or epidemiological evidence, nor has a diagnostic test been performed. Hence the distribution of suspected/probable/confirmed will depend on the pathogen characteristics, outbreak-specific definitions, and resources available.

The line list output from the {simulist} simulation contains a column (`case_type`) with the type of case. 

{simulist} can simulate varying probabilities of each case being suspected, probable or confirmed. By default the `sim_linelist()` function uses probabilities of `suspected = 0.2`, `probable = 0.3` and `confirmed = 0.5`. 

```{r, sim-linelist-default-case-type}
linelist <- sim_linelist(
  mean_contacts = 2,
  contact_interval = contact_interval,
  prob_infect = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death
)
head(linelist)
```

To alter these probabilities, supply a named vector to the `sim_linelist()` argument `case_type_probs`. The vector should contain three numbers, with the names `suspected`, `probable` and `confirmed`, with the numbers summing to one. Here we change the values to simulate an outbreak in which the proportion of cases confirmed by laboratory testing is high.

```{r, sim-linelist-mod-case-type}
linelist <- sim_linelist(
  mean_contacts = 2,
  contact_interval = contact_interval,
  prob_infect = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death,
  case_type_probs = c(suspected = 0.05, probable = 0.05, confirmed = 0.9)
)
head(linelist)
```

It is also possible to set one of these categories to `1`, in which case every case will be of that type. 

The way {simulist} assigns case types is by pasting case types onto existing case data. Thus, it could be viewed that the true underlying data is that all cases in the simulation are confirmed, but that there is a lack of information in some cases. There are no cases in the output line list that are incorrectly attributed as probable or suspected that have not been infected. That is to say, all individuals in the line list, whatever their case type, are infected during the outbreak.

## Conditioning simulation on outbreak size

The reproduction number has a strong influence on the size of an outbreak. For {simulist}, the reproduction number is determined by the mean number of contacts and the probability of infection. However, the {simulist} package generates line list data using a stochastic algorithm, so even when $R < 1$ it can produce a substantial outbreak by chance, or an $R >> 1$ will sometimes not produce a vast epidemic in one simulation (i.e. one replicate) due to the stochasticity. 

When requiring a minimum outbreak size we can specify the `min_outbreak_size` argument in `sim_linelist()`. By default this is set to 10. This means that the simulation will not return a line list until the conditioning has been met. In other words, the simulation will resimulate a branching process model until an outbreak infects at least 10 people.

When requiring a line list that represents a large outbreak, such as the COVID-19 outbreak, setting the `min_outbreak_size` to a larger number guarantees a line list of at least that size. Here we simulate a line list requiring at least 250 cases.

```{r, sim-large-linelist}
linelist <- sim_linelist(
  mean_contacts = 2,
  contact_interval = contact_interval,
  prob_infect = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death,
  min_outbreak_size = 250
)
head(linelist)
```

The amount of time the simulation takes can be determined by the mean number of contacts (`mean_contacts`), the probability of infection (`prob_infect`) and the minimum outbreak size (`min_outbreak_size`). If the `min_outbreak_size` is large, for example hundreds or thousands of cases, and the mean number of contacts and probability of infection mean the reproduction number is below one, it will take many branching process simulations until finding one that produces a sufficiently large epidemic.

## Anonymous line list

By default `sim_linelist()` provides the name of each individual in the line list. If an anonymised line list is required the `add_names` argument of `sim_linelist()` can be set to `FALSE`. 

```{r sim-anon-linelist}
linelist <- sim_linelist(
  mean_contacts = 2,
  contact_interval = contact_interval,
  prob_infect = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death,
  add_names = FALSE
)
head(linelist)
```

The names used in the line list are produced at random by the [{randomNames}](https://CRAN.R-project.org/package=randomNames) R package. Therefore, even when `add_names = TRUE` there is no personal data of real individuals being produced or shared.

## Population age

For an overview of how a line list can be simulated with a uniform or structured population age distribution see the [vignette dedicated to this topic](age-struct-pop.html).

## Age-stratified hospitalisation and death risks

For an overview of how a line list can be simulated with age-stratified (or age-dependent) hospitalisation and death risks see the [vignette dedicated to this topic](age-strat-risks.html).

## Simulate contacts table

To simulate a contacts table, the `sim_contacts()` function can be used. This requires the same arguments as `sim_linelist()`, but does not require the onset-to-hospitalisation delay and onset-to-death delays.

```{r, sim-contacts}
contacts <- sim_contacts(
  mean_contacts = 2, 
  contact_interval = contact_interval, 
  prob_infect = 0.5)
head(contacts)
```

## Simulate both line list and contacts table

To produce both a line list and a contacts table for the same outbreak, the `sim_linelist()` and `sim_contacts()` cannot be used separately due to the stochastic algorithm, meaning the data in the line list will be discordant with the contacts table. 

In order to simulate a line list and a contacts table of the same outbreak the `sim_outbreak()` function is required. This will simulate a single outbreak and return a line list and a contacts table. The inputs of `sim_outbreak()` are the same as the inputs required for `sim_linelist()`.

```{r, sim-outbreak}
outbreak <- sim_outbreak(
  mean_contacts = 2, 
  contact_interval = contact_interval, 
  prob_infect = 0.5, 
  onset_to_hosp = onset_to_hosp, 
  onset_to_death = onset_to_death
)
head(outbreak$linelist)
head(outbreak$contacts)
```

`sim_outbreak()` has the same features as `sim_linelist()` and `sim_contacts()`, this includes simulating with age-stratified risks of hospitalisation and death, the probability of case types or contact tracing status can be modified.

## Using functions for distributions instead of `<epidist>`

It is possible to use an [anonymous function](https://en.wikipedia.org/wiki/Anonymous_function) instead of an `<epidist>` object when specifying the parameters of the delay and contact distributions. We recommend using the `<epidist>` objects but here outline the alternative approach.

```{r, sim-outbreak-anon-func}
outbreak <- sim_outbreak(
  mean_contacts = 2,
  contact_interval = function(x) rgamma(n = x, shape = 2, scale = 2),
  prob_infect = 0.5,
  onset_to_hosp = function(x) rlnorm(n = x, meanlog = 1.5, sdlog = 0.5),
  onset_to_death = function(x) rweibull(n = x, shape = 0.5, scale = 0.2)
)
head(outbreak$linelist)
head(outbreak$contacts)
```

The same approach of using anonymous functions can be used in `sim_linelist()` and `sim_contacts()`.
