---
title: "Visualising simulated data"
output: 
  bookdown::html_vignette2:
    code_folding: show
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Visualising simulated data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette gives an overview of the ways to plot the line list data output from the `sim_linelist()` function.

Plotting can be useful to identify certain transmission dynamics and patterns in the simulated data, or just to check that the simulated data looks as expected given how the simulation was parameterised.

```{r setup}
library(simulist)
library(epiparameter)
library(incidence2)
library(epicontacts)
```

First we load the required delay distributions using the {epiparameter} package.

```{r, read-delay-dists}
serial_interval <- epidist(
  disease = "COVID-19",
  epi_dist = "serial interval",
  prob_distribution = "gamma",
  prob_distribution_params = c(shape = 1, scale = 1)
)

# get onset to hospital admission from {epiparameter} database
onset_to_hosp <- epidist_db(
  disease = "COVID-19",
  epi_dist = "onset to hospitalisation",
  single_epidist = TRUE
)

# get onset to death from {epiparameter} database
onset_to_death <- epidist_db(
  disease = "COVID-19",
  epi_dist = "onset to death",
  single_epidist = TRUE
)
```

Setting the seed ensures we have the same output each time the vignette is rendered. When using {simulist}, setting the seed is not required unless you need to simulate the same line list multiple times.

```{r, set-seed}
set.seed(1)
```

Using a simple line list simulation with the factory default settings:

```{r sim-linelist} 
linelist <- sim_linelist(
  R = 1.3,
  serial_interval = serial_interval,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death
)
```


```{r num-cases, echo=FALSE}
message("This line list contains ", nrow(linelist), " cases.")
```

## Visualising incidence of onset, hospitalisation and death

::: {.alert .alert-info}

This section of the vignette is heavily based upon examples given in the [Get Started vignette in the {incidence2} package](https://www.reconverse.org/incidence2/articles/incidence2.html). It is highly recommended to read the documentation supplied in the [{incidence2} package](https://CRAN.R-project.org/package=incidence2) to explore the full range of functionality.

:::

To visualise the number of cases with onset on a particular day, the {incidence2} package, and its dedicated class (`<incidence2>`) are used for handling and plotting this data.

Currently {simulist} outputs dates that are not rounded to the nearest day, i.e. it can be half way through a day. This is not obvious as R prints dates to the nearest day by default, and only by removing the date class (using `unclass()`) can you see the decimals (as R stores dates internally as the number of days since 1970-01-01).

***Note*** storing dates as precise doubles and not as integer days may change in the near future.

The `interval = "daily"` is required as {incidence2} requires rounded dates to aggregate cases per unit time and specifying the `interval` will do this automatically for us.

```{r create-incidence}
# create incidence object
daily <- incidence(x = linelist, date_index = "onset_date", interval = "daily")
```

It is possible that not every date had the onset of symptoms, resulting in some dates missing entries. This is taken care of by the `complete_dates()` function in {incidence2}.

```{r, complete-dates, fig.cap="Daily incidence of cases from symptom onset including days with zero cases.", fig.width = 8, fig.height = 5}
# impute for days without cases
daily <- complete_dates(daily)
plot(daily)
```

Alternatively, incidence can be plotting weekly:

```{r plot-weekly, fig.cap="Weekly incidence of cases from symptom onset", fig.width = 8, fig.height = 5}
weekly <- incidence(linelist, date_index = "onset_date", interval = "isoweek")
plot(weekly)
```

In order to check differences between a group in the line list data, for example gender, the `<incidence2>` data object can be recreated, specifying which columns to group by.

```{r, group-by-gender, fig.cap="Weekly incidence of cases from symptom onset facetted by gender.", fig.width = 8, fig.height = 5}
weekly <- incidence(
  linelist,
  date_index = "onset_date",
  interval = "isoweek",
  groups = "gender"
)
plot(weekly)
```
 
To visualise the onset, hospitalisation and death incidence in the same plot they can be jointly specified to the `date_index` argument of `incidence2::incidence()`. 
 
```{r, plot-onset-hospitalisation, fig.cap="Daily incidence of cases from symptom onset and incidence of hospitalisations and deaths.", fig.width = 8, fig.height = 5}
daily <- incidence(
  linelist,
  date_index = c(
    onset = "onset_date",
    hospitalisation = "hospitalisation_date",
    death = "death_date"
  ),
  interval = "daily",
  groups = "gender"
)
daily <- complete_dates(daily)
plot(daily)
```

## Demographic data

Please see the [Age structured population vignette](age-struct-pop.html) for examples of how to plot the distribution of ages within a line list data set, including age pyramids. 

The plotting code in vignettes is hidden by default, click the Code button with arrow to reveal the plotting code.

## Visualising contact data

::: {.alert .alert-info}

This section of the vignette is based upon examples from the [{epicontacts} R package documentation](https://www.repidemicsconsortium.org/epicontacts/index.html) and the examples provided in the [The Epidemiological R Handbook chapter on transmission chains](https://epirhandbook.com/en/transmission-chains.html). We recommend going to the documentation of the {epicontacts} R package to see the all plotting and data wrangling functionality.

:::

Just as we utilised the `<incidence2>` class from the {incidence2} package to handle and plot incidence data, we are going to use the `<epicontacts>` class from the {epicontacts} R package to handle and plot epidemiological contact data.

::: {.alert .alert-success}

The benefit of using {epicontacts} is the same as {incidence2}, in the fact that a default plotting method is supplied by the package.

_Advanced_

Additionally, for {epicontacts} there is access to network plotting from Javascript libraries via the {visnetwork} and {threejs} R packages.

:::

The {epicontacts} function `make_epicontacts()` requires both the line list and contact table, so we will run the `sim_outbreak()` function to produce both. We will use the same epidemiological delay distributions that we used to simulate a line list above. We need an extra distribution to simulate an outbreak, a contact distribution (see the documentation `?sim_outbreak` for more detail). We create this distribution using the {epiparameter} package.

```{r, create-contact-dist}
contact_distribution <- epiparameter::epidist(
  disease = "COVID-19",
  epi_dist = "contact_distribution",
  prob_distribution = "pois",
  prob_distribution_params = c(l = 5)
)
```

Now we can simulate an outbreak:

```{r, sim-outbreak}
set.seed(1)
outbreak <- sim_outbreak(
  R = 1.1,
  serial_interval = serial_interval,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death,
  contact_distribution = contact_distribution
)
head(outbreak$linelist)
head(outbreak$contacts)
```

Using the line list and contacts data simulated we can create the `<epicontacts>` object.

```{r, create-epicontacts}
epicontacts <- make_epicontacts(
  linelist = outbreak$linelist, 
  contacts = outbreak$contacts, 
  id = "case_name", 
  from = "from", 
  to = "to", 
  directed = TRUE
)
```

The `<epicontacts>` object comes with a custom printing feature to see the data.

```{r, print-epicontacts}
epicontacts
```

To plot the contact network we can use the plotting method that is supplied by epicontacts and will be automatically recognised if the {epicontacts} package is loaded (as done above with `library(epicontacts)`).

```{r, plot-epicontacts}
plot(epicontacts)
```

There is also the option to plot the contacts network in 3D using the `epicontacts::graph3D()`. 

## Visualising other line list information

If there are other aspects of line list data that can be plotted and you would like to them added to this vignette please make an [issue](https://github.com/epiverse-trace/simulist/issues) or [pull request](https://github.com/epiverse-trace/simulist/pulls).
