---
title: "Wrangling simulated outbreak data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wrangling simulated outbreak data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The {simulist} R package can generate line list data (`sim_linelist()`), contact tracing data (`sim_contacts()`), or both (`sim_outbreak()`). By default the line list produced by `sim_linelist()` and `sim_outbreak()` contains 12 columns. Some amount of post-simulation data wrangling may be needed to use the simulated epidemiological case data to certain applications. This vignette demonstrates some common data wrangling tasks that may be performed on simulated line list or contact tracing data.

```{r setup}
library(simulist)
library(epiparameter)
library(dplyr)
```

This vignette provides data wrangling examples using both functions available in the R language (commonly called "base R") as well as using [tidyverse R packages](https://www.tidyverse.org/), which are commonly applied to data science tasks in R. The tidyverse examples are shown by default, but select the "Base R" tab to see the equivalent functionality using base R. There are many other tools for wrangling data in R which are not covered by this vignette (e.g. [{data.table}](https://rdatatable.gitlab.io/data.table/)).

::: {.alert .alert-info}
See these great resources for more information on general data wrangling in R:

* [R for Data Science by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund](https://r4ds.hadley.nz/)
* [{dplyr} R package](https://dplyr.tidyverse.org/)
* [{tidyr} R package](https://github.com/tidyverse/tidyr)
* [Wrangling data frames chapter in An Introduction to R by Alex Douglas, Deon Roos, Francesca Mancini, Ana Couto & David Lusseau](https://intro2r.com/wrangling-data-frames.html)
:::

## Simulate an outbreak

To simulate an outbreak we will use the `sim_outbreak()` function from the {simulist} R package.

::: {.alert .alert-info}
If you are unfamiliar with the {simulist} package or the `sim_outbreak()` function [Get Started vignette](simulist.html) is a great place to start.
:::

First we load in some data that is required for the outbreak simulation. Data on epidemiological parameters and distributions are read from the {epiparameter} R package.

```{r read-epidist}
# create contact distribution (not available from {epiparameter} database)
contact_distribution <- epiparameter(
  disease = "COVID-19",
  epi_name = "contact distribution",
  prob_distribution = create_prob_distribution(
    prob_distribution = "pois",
    prob_distribution_params = c(mean = 2)
  )
)

# create infectious period (not available from {epiparameter} database)
infectious_period <- epiparameter(
  disease = "COVID-19",
  epi_name = "infectious period",
  prob_distribution = create_prob_distribution(
    prob_distribution = "gamma",
    prob_distribution_params = c(shape = 1, scale = 1)
  )
)

# get onset to hospital admission from {epiparameter} database
onset_to_hosp <- epiparameter_db(
  disease = "COVID-19",
  epi_name = "onset to hospitalisation",
  single_epiparameter = TRUE
)

# get onset to death from {epiparameter} database
onset_to_death <- epiparameter_db(
  disease = "COVID-19",
  epi_name = "onset to death",
  single_epiparameter = TRUE
)
```

The seed is set to ensure the output of the vignette is consistent. When using {simulist}, setting the seed is not required unless you need to simulate the same line list multiple times.

```{r, set-seed}
set.seed(123)
```

```{r, sim-outbreak}
outbreak <- sim_outbreak(
  contact_distribution = contact_distribution,
  infectious_period = infectious_period,
  prob_infection = 0.5,
  onset_to_hosp = onset_to_hosp,
  onset_to_death = onset_to_death
)
linelist <- outbreak$linelist
contacts <- outbreak$contacts
```

## Censoring dates

The date event columns in simulated line lists are stored to double point precision, meaning they are the exact event times. It is unusual to not store `<Date>` objects as integers, as explained in `?Dates`, and the `print()` function for `<Date>`s does not show that they may be part way through a day.

Here we show this by printing the date of symptom onset for the simulated data, and then unclass it to show how it is stored internally.

```{r}
linelist$date_onset
unclass(linelist$date_onset)
```

The `censor_linelist()` function can be used after `sim_linelist()` to censor the event dates to a given precision. Here we show censoring the dates to daily and weekly intervals. The daily censoring dates will look that same as before, but the dates will have any value after the decimal point set to zero. The weekly censored dates will be printed differently.

```{r}
daily_cens_linelist <- censor_linelist(linelist, interval = "daily")
head(daily_cens_linelist)

weekly_cens_linelist <- censor_linelist(linelist, interval = "weekly")
head(weekly_cens_linelist)
```

See `?censor_linelist()` for more information on how to use this function.

By using `censor_linelist()` it avoids common mistakes when working with `<Date>` objects. For example, rounding a date that is over half way through a day will mistakenly result in the next day. Using `censor_linelist()` avoids this and other common mistakes.

```{r}
linelist$date_onset
round(linelist$date_onset)
daily_cens_linelist$date_onset
```

## Removing a line list column {.tabset}

Not every column in the simulated line list may be required for the use case at hand. In this example we will remove the `$ct_value` column. For instance, if we wanted to simulate an outbreak for which no laboratory testing (e.g Polymerase chain reaction, PCR, testing) was available and thus a Cycle threshold (Ct) value would not be known for confirmed cases.

### Tidyverse

```{r, rm-ct-col-tidyverse}
# remove column by name
linelist %>% # nolint one_call_pipe_linter
  select(!ct_value)
```

### Base R

```{r, rm-ct-col-base}
# remove column by numeric column indexing
# ct_value is column 12 (the last column)
linelist[, -12]

# remove column by column name
linelist[, colnames(linelist) != "ct_value"]

# remove column by assigning it to NULL
linelist$ct_value <- NULL
linelist
```

## {-}
